NanoMaLy

### 2015-11-18 paper/pldi16-submission/semantics.tex

\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_0$, $e'$ is $e_j$ and $\forall 0 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of functions:
called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
that ``dynamically'' perform type-checking and hole-filling
respectively.

\paragraph{Narrowing Types} The function $\force{e}{t}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t}
by yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t},
or yields the stuck term if no such refinement is possible:
%
$$
\begin{array}{lcl}
\forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
\force{v}{\ehole{}}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

\paragraph{Generating Values} The function $\gen{t}$ takes as input a type
\texttt{t} and returns \RJ{HEREHERE}

$$
\begin{array}{lcll}
\gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b & \quad \text{non-deterministic} \\
\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
\end{array}
$$

%

Note that each primitive reduction step -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation.
%
Additionally, beta-reduction \emph{does not} type-check its argument, it
only ensures that the value being applied is a function.
%
\begin{thm}
\label{thm:all-reduce}
  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
  \ES{do we really need to state this, or is it obvious?}
\end{thm}

%
% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}
%
\subsection{Soundness}
\label{sec:soundness}
We now show that our evaluation relation conservatively instantiates
holes, \ie if, given a function $f$, we report that
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.

We can think of the evaluation of \eapp{f}{\ehole{}} as computing a
\emph{partial type} -- a type that may contain holes -- for $f$.
%
We can extract this type from the result of evaluation as follows.
%
\begin{defn}
\label{defn:partial-type}
  If \stepi{i}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su}, then the $i$th
  \emph{partial type} of f, written \ptype{i}{f},
  is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.

  We will omit the subscript when we wish to refer to the final partial
  type.
\end{defn}
%
The \typeof{} function is an approximation of the type of an expression.
\begin{defn}
\label{def:typeof}
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
  \end{array}
  \]
\end{defn}
%
We also define a compatibility relation between types.
%
\begin{defn}
\label{defn:type-compat}
  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
  \ES{we're abusing the \su notation here..}
\end{defn}
%
Given these two definitions, we show that each evaluation step
refines the partial type of $f$.
%
\begin{lem}
\label{lem:refine-partial}
  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
  have been called at step $k+1$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at step
  $k+1$.
  %
  An examination of the rules shows that only place this happens is
  in the first case of \forcesym.
\end{proof}
%
Finally, we show that any value that is not compatible with the $k$th
partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
steps.
%
\begin{lem}
\label{lem:k-stuck}
  For any \hastype{v}{t} that is not compatible with
  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$.

  Let $\tfun{s_k}{\_} = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
  \begin{description}
  \item[Case \tincompat{t}{s_k}:]
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}
%
Now we can prove our soundness theorem.
%
\begin{thm}
\label{thm:soundness}
  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}
\begin{proof}
  Let $\tfun{s}{\_} = \ptype{}{f}$.
  \begin{description}
  \item[Case \tcompat{s}{t}:] Our witness is already valid.
  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
    use \emph{any} \hastype{v}{t} as a witness for $f$.
  \end{description}
\end{proof}

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
We have shown how to find a witness for a function of a single argument,
but in a language with higher-order functions and currying it may not be
clear \emph{syntactically} how many arguments a function takes.
%
Thus, we wrap our operational semantics for \lang in a search loop that
supplies an increasing number of arguments until the function returns a
value that is not a lambda.

The top-level search loop takes as input an open program -- a sequence
of binders -- and searches for an expression that closes the input
program and gets stuck. Concretely, given an input
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
we will search for an expression of the form:
%
\begin{code}
  fn v1 ... vn
\end{code}
%
Figure~\ref{fig:expression-api} describes a small API for manipulating
and evaluating \lang expressions, which we will use to define our search
algorithm.
%
The bulk of the search is performed by @eval@, which
implements our operational semantics from \S\ref{sec:semantics}.
%
The operational semantics is non-deterministic due to \gensym,
thus @eval@ returns a list of possible results.
%
\begin{figure}[t]
  \centering
  \begin{mcode}
  -- transitive small-step reduction,
  -- returning a list of results
  eval :: ($e$, $\su$) -> [($v$, $\su$)]

  -- manipulating expressions
  subst   :: $\su$ -> $v$ -> $v$
  mkApps  :: $e$ -> [$e$] -> $e$
  mkLets  :: [($x$, $e$)] -> $e$ -> $e$
  isStuck :: $v$ -> Bool
  \end{mcode}
  \caption{Expression API}
  \label{fig:expression-api}
\end{figure}
%
We also define a few helper functions for manipulating expressions:
\begin{itemize}
\item @subst@ applies a substitution of holes to a value,
\item @mkApps@ creates a nested sequence of applications in the usual
  left-associative style,
\item @mkLets@ takes a list of binders and a body expression, and
  creates a sequence of nested let-binders, and
\item @isStuck@ tests whether a value is the \stuck term.
\end{itemize}
%
\begin{figure*}[t]
  \centering
  \begin{mcode}
  check :: [($x$, $e$)] -> Result
  check bnds =
    -- (2) search for a witness
    case find (isStuck . fst) results of
      Nothing      -> Safe
      Just (_, su) -> Unsafe (mkApps f (subst su args))
    where
      (args, results) = loop []
      f               = snd (last bnds)
      build args      = mkLets bnds (mkApps f args)

      -- (1) find the correct number of arguments
      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
      loop args = case eval (build args, []) of
        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
        results      -> (args, results)
  \end{mcode}
  \caption{Implementing our counter-example search in terms of the
    operational semantics.
    \ES{1 and 2 are backwards, ugh...}
    \ES{should address case where output types of successive runs dont match}
  }
  \label{fig:search-algo}
\end{figure*}
%
Figure~\ref{fig:search-algo} summarizes the overall implementation of
our search for witnesses, which takes as input a sequences of binders
and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
where $e$ is a term that causes the input program to get stuck.
%
The search is split into two phases: (1) we supply an increasing number
of arguments until we find a saturated application, and (2) we search
through the list of results from the non-deterministic evaluator for a
witness.

% Algorithm:
%   Input: ML Program (let f1 = e1 in let f2 = e2 in ... e)
%   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck

>>>>

\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%

The evaluation relation is parameterized by a pair of functions:
called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
that ``dynamically'' perform type-checking and hole-filling
respectively.

\paragraph{Narrowing Types} The procedure $\force{e}{t}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t}
by yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t},
or yields the stuck term if no such refinement is possible:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
\force{v}{\ehole{}}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type \texttt{t} and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole{}$) yield fresh, unconstrained values
(denoted by $\ehole{i}$).

$$
\begin{array}{lcll}
% \gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b & \quad \text{non-deterministic} \\
\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
\end{array}
$$

\paragraph{Steps and Traces}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
%
We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
%
We write \steps{e}{\vsu}{e'}{\vsu'} if \steps{e}{\vsu}{e'}{\vsu'}
for some trace $\trace$.

\paragraph{Primitive Reductions}
Primitive reduction steps -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

%% \begin{thm}
%% \label{thm:all-reduce}
%%   Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
%% \ES{do we really need to state this, or is it obvious?}
%% \end{thm}

% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}
%
\subsection{Generality}\label{sec:soundness}

Recall that a key technical challenge in generating witnesses is
that we have no (static) type information to rely upon. Thus, we
must avoid the trap of generating \emph{spurious} witnesses that
arise from picking irrelevant values, when instead there exist
perfectly good values of a \emph{different} type, under which
the program would not go wrong.

\paragraph{Witness Generality} We can now define
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
possible input type $t$, we can find a value $\hastype{v}{t}$
such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}

We need to develop some machinery in order to prove this theorem.
First, we show how our evaluation rules encode a dynamic form of
type inference, and next, we show that the types inferred via
evaluation are indeed maximally general.

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
  \end{array}
  \]

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
then the \emph{partial type of} $f$ upto $\trace$, written \ptype{\trace}{f},
is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

\paragraph{Type Compatibility}
Two types are compatible if one of them can be obtained by
suitably filling the holes of the other (\ie by suitably ``instantiating''
the type variables appearing in the other). Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ t = \subst{\tsu}{s} \lor s = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
and
$\trace' = \trace, \steps{e}{\vsu}{e'}{\vsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\steps{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at the last step.
  %
  An examination of the rules shows that only place this happens is
  in the first case of \forcesym.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
We show that \emph{any} value that is \emph{incompatible} with
the partial type upto $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of the
trace $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     not \tcompat{t}{\ptype{\trace}{f}} and
     \hastype{v}{t}
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}

\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_k}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
  \begin{description}
  \item[Case \tincompat{t}{s_k}:]
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}

\paragraph{Proof of Theorem~\ref{thm:soundness}}
%
Now we can prove our soundness theorem. Intuitively, suppose $\trace$
witnesses that $f$ gets stuck, and let $t$ be the partial type of $f$
upto $\trace$. We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is compatible with the
partial type upto $\trace$.
%
\begin{description}
  \item[Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  \item[Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
\end{description}

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
So far, we have seen how a trace leading to a stuck configuration yields
a general witness demonstrating that the program is ill-typed (\ie\ goes
wrong for all inputs.)
In particular, we have shown how to non-deterministically find a witnesses
for a function of a \emph{single} argument.

In order to convert the semantics into a \emph{procedure} for finding
witnesses, we must address two challenges.
%
First, we must resolve the non-determinism introduced by \gensym.
%
Second, in the presence of higher-order functions and currying,
we must determine how many concrete values to generate to make
execution go wrong (as we cannot rely upon static typing to
provide this information.)

The witness generation procedure @genWitness@ is formalized in
Figure~\ref{fig:algo-gen-witness}.
%
Next, we describe what it takes as input and output, and how it
addresses the above challenges to search the space of possible
executions for general type error witnesses.

\paragraph{Inputs and Outputs}
%
Obviously, the problem of generating inputs is undecidable in general.
%
The witness generation procedure takes two inputs.
%
First, a search bound @n@ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
denoting the list of \emph{top-level definitions}
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
The witness generation procedure returns as output a list of (general)
witness expressions $e$ each of which is of the form @fn v1 ... vn@.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.
%
In the sequel, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: ($e$, $\su$) -> [($v$, $\su$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call @eval($e$, $\vsu$)@
returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
that $\steps{e}{\vsu}{e'}{\vsu'}$.
%
We determine whether a given value is stuck by using the predicate
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that
@genArgs [(f1,e1),...,(fn, en)]@ yields a list of holes @[v1,...,vk]@
such that @fn v1 ... vk@ \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes until
@eval@uating the closed term yields a non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking @eval@.
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.

\begin{figure}[ht]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
               _        -> vs

close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$)
close bs vs = (e, $\emptysu$)
  where
    e       = mkLets bs (witness bs vs)

witness :: [($x$, $e$)] -> [$v$] -> $e$
witness bs vs = mkApps target vs
  where
    target    = snd (last bs)
\end{mcode}
\caption{Find number of arguments for target function}
\label{fig:algo-gen-args}
\end{figure}

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in procedure @genWitness@
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ which demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:

\begin{thm}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k@ $[(f_1, e_1), \ldots, (f_n, e_n)]$ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  \steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}

\begin{figure}[ht]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []      -- (1)
   $e$    = close defs vs        -- (2)
   res = take n (eval e)      -- (3)
   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

%% \begin{figure}[t]
  %% \centering
  %% \begin{mcode}
  %% -- transitive small-step reduction, returning a list of results
  %% eval :: ($e$, $\su$) -> [($v$, $\su$)]
%%
  %% -- | is a value stuck?
  %% isStuck :: $v$ -> Bool
%%
  %% mkApps  :: $e$ -> [$e$] -> $e$
  %% mkLets  :: [($x$, $e$)] -> $e$ -> $e$
  %% \end{mcode}
  %% \caption{Expression API}
  %% \label{fig:expression-api}
%% \end{figure}
%% %
%% We also define a few helper functions for manipulating expressions:
%% \begin{itemize}
%% \item @subst@ applies a substitution of holes to a value,
%% \item @mkApps@ creates a nested sequence of applications in the usual
  %% left-associative style,
%% \item @mkLets@ takes a list of binders and a body expression, and
  %% creates a sequence of nested let-binders, and
%% \item @isStuck@ tests whether a value is the \stuck term.
%% \end{itemize}

%%%  \begin{figure*}[t]
  %%%  \centering
  %%%  \begin{mcode}
  %%%  check :: [($x$, $e$)] -> Result
  %%%  check bnds =
    %%%  -- (2) search for a witness
    %%%  case find (isStuck . fst) results of
      %%%  Nothing      -> Safe
      %%%  Just (_, su) -> Unsafe (mkApps f (subst su args))
    %%%  where
      %%%  (args, results) = loop []
      %%%  f               = snd (last bnds)
      %%%  build args      = mkLets bnds (mkApps f args)
%%%
      %%%  -- (1) find the correct number of arguments
      %%%  loop :: [$v$] -> ([$v$], [($v$, $\su$)])
      %%%  loop args = case eval (build args, []) of
        %%%  ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
        %%%  results      -> (args, results)
  %%%  \end{mcode}
  %%%  \caption{A procedure for generating witnesses}
    %%%  \ES{should address case where output types of successive runs dont match}
  %%%  }
  %%%  \label{fig:search-algo}
%%%  \end{figure*}
%
